Multilevel selection in causal models: the non-ergodicity of evolutionary and probability theories as the general driver of the emergence of cooperation and specialization.

### 2022-01-21 article.tex
Version 0.0.0

%opening
\title{Multilevel selection in causal models: the non-ergodicity of evolutionary and probabilistic theories as the general driver of the emergence of cooperation and specialization. }
\author{Gustavo Landfried}

@@

\section{\en{Introduction}\es{Introducción}}

@@

\subsection{\en{Isomorphism between evolutionary and probabilistic theories}\es{Isomorfismo entre las teorías de la evolución y la probabilidad}}

@@

\en{In this section we will analyze whether there is indeed any evolutionary advantage of cooperation and specialization in the presence of defection. }%
\es{En esta sección analizaremos si efectivamente existe alguna ventaja evolutiva de la cooperación y la especialización en presencia de deserción. }%
%
\en{To do so, we will extend the causal model proposed in the methodology section, by simply incorporating the cooperative behavior presented in the introduction section, including a defector behavior (who receives the benefit of cooperation without contributing to it). }%
\es{Para ello, extenderemos el modelo causal propuesto en la sección metodología, incorporando simplemente el comportamiento cooperativo presentado en la sección introducción, incluyendo un comportamiento desertor (que recibe el beneficio de la cooperación sin aportar al mismo). }%
%
\en{This modification alone has a fundamental evolutionary consequence: the formation of groups and thus the emergence of a new evolutionary entity. }%
\es{Esta sola modificación tiene una consecuencia evolutiva fundamental: la formación de grupos y por lo tanto la aparición de una nueva entidad evolutiva. }
%
\en{Because, while defecting individuals may have a relative advantage over cooperating individuals, groups may be favored by mutual cooperation. }%
\es{Porque, si bien los individuos desertores puede tener una ventaja relativa sobre los individuos cooperadores, los grupos pueden verse favorecidos por la mututa cooperación. }%
%
\en{Darwin has already discussed this topic. }
\es{Darwin ya discutió sobre este tema. }
%
\begin{quotation}
It must not be forgotten that although a high standard of morality gives but a slight or no advantage to each individual man and his children over the other men of the same tribe, yet that an increase in the number of well-endowed men and an advancement in the standard of morality will certainly given an immense advantage to one tribe over another.
\end{quotation}
%
\en{The incorporation of cooperation in the probabilistic causal model induces the formation of groups and therefore a type of group-level selection that acts in parallel to individual-level selection. }%
\es{La incorporación de la cooperación en el modelo causal probabilísitico induce la formación de grupos y por lo tanto un tipo de selección a nivel grupal que actúa de forma paralela a la selección de nivel individual. }%
%
\en{The joint result of both selection processes, known as multilevel selection, will be the property of interest. }%
\es{El resultado de ambos procesos de selección, conocido como selección multinivel, será la propiedad de interés. }%

\subsection{Modelo causal probabilísitico.}

@@

\en{With $N$ individuals, there are $N+1$ types of possible regions, from $n=0$ (all defectors) to $n=N$ (all cooperators). }%
\es{Con $N$ individuos, hay $N+1$ tipos de regiones posibles, desde $n=0$ (todos desertores) hasta $n=N$ (todos cooperadores). }%

@@

\en{and the strategy $\Ee_i$ they use to allocate resources $f_1(\Ee,\Aa)$ (equation \ref{eq:familia_de_aptitudes}). }%
\es{y la estrategia $\Ee_i$ que utilizan para alocar recursos $f_1(\Ee,\Aa)$ (ecuación~\ref{eq:familia_de_aptitudes}). }%
\en{A priori, there is no preference for defector or cooperative social behavior. }%
\es{A priori, no hay preferencia por el comportamiento social desertor o cooperador. }%
\en{Then, by combinatorics, it will be more likely a priori that individuals will inhabit mixed regions. }%
\es{Entonces, por simple combinatoria, será más probable a priori que los individuos habiten regiones mixtas. }%
P(i) = \frac{1}{N} \text{Binomial}(\texttt{region}(i),N,0.5) 
\en{where the binomial distribution gives more weight to individuals from mixed regions, the parameter $0.5$ indicates a uniform prior between cooperating and defecting social behaviors, and the factor $\frac{1}{N}$ indicates that all individuals in the same region start with the same initial resources. }%
\es{donde la distribución binomial otorga más peso a los individuos de las regiones mixtas, el parámetro $0.5$ indica un prior uniforme entre los comportamientos sociales cooperador y desertor, y el factor $\frac{1}{N}$ indica que todos los individuos de la misma región empiezan con los mismos recursos iniciales. }%
\en{As we have seen in the methodology section, the probability of the environmental states is, }%
\es{Como hemos visto en la sección metodología, la probabilidad de los estados ambientales es, }%

@@

\en{Finally, the groups are constituted by the members of each region, $P(g|i)=\mathbb{I}(\texttt{region}(i) = g)$, the indexing function which is $1$ when individuals belong to region $g$ and $0$ otherwise. }%

@@

\en{The evolutionary problem we are interested in is the selection of cooperative individuals given the environment within each group (level 1), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)$, the selection of groups given the environment (level 2), $P(g^T|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$,  and selection of cooperative individuals given the environment for all groups (multilevel), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$. }%
\es{El problema evolutivo que nos interesa es la selección de los individuos cooperadores dado el ambiente dentro de cada grupo (nivel 1), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)$, la selección de los grupos dado el ambiente (nivel 2), $P(g^T|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$, y la selección de los individuos cooperadores dado el ambiente integrando todos los grupos (multinivel), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$. }%
%
\en{The multilevel selection is obtained by integrating the product of level 1 and 2 selections, }%
\es{La selección multinivel se obtiene integrando el producto de las selecciones de nivel 1 y 2, }%
%
\begin{equation}\label{eq:posterior_multinivel}
\underbrace{P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})}_{\text{\en{Multilevel selection}\es{Selección multinivel}}} = \sum_{g=0}^N \underbrace{P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)}_{\text{\en{Level 1 selection}\es{Selección de nivel 1}}} \cdot \underbrace{P(g^T|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})}_{\text{\en{Level 2 selection}\es{Selección de nivel 2}}}
\end{equation}
%
\en{where the posterior of level 1 selection is obtained by integrating the probability of all cooperating individuals that are part of the group $g$, }%
\es{donde el posterior de la selección de nivel 1 se obtiene integrando la probabilidad de todos los individuos cooperadores que forman parte del grupo $g$, }%
%
\begin{equation}\label{eq:posterior_nivel_1}
\begin{split}
P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g) &= \sum_{j=1}^M P(I^T=j|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)\mathbb{I}(\texttt{coop}(j))
\end{split}
\end{equation}

\subsection{Probabilidades marginales}

\en{To compute the posteriors it will be sufficient to compute the marginal probabilities, since both expressions are proportional. }%
\es{Para calcular los posteriors va a ser suficiente con calcular las probabilidades marginales, pues ambas expresiones son proporcionales. }%

>>>>

%opening
\title{Multilevel selection in causal models: the non-ergodicity of evolutionary and probability theories as the general driver of the emergence of cooperation and specialization. }
\author{Gustavo Landfried}

@@

\section{\en{Introduction}\es{Introducción}} \label{sec:introduction}

@@

\subsection{\en{Isomorphism between evolutionary and probability theories}\es{Isomorfismo entre las teorías de la evolución y la probabilidad}}

@@

% \en{In this section we will analyze whether there is indeed any evolutionary advantage of cooperation and specialization in the presence of defection. }%
% \es{En esta sección analizaremos si efectivamente existe alguna ventaja evolutiva de la cooperación y la especialización en presencia de deserción. }%
% %
\en{To analyze whether there is any evolutionary advantage in favor of cooperation and specialization, in this section we will extend the causal model proposed in the \ref{sec:methodology} section by incorporating unconditionally cooperative behaviors, such as the one presented in the \ref{sec:introduction} section, including defecting behaviors (which receive the benefit of cooperation without contributing to it). }%
\es{Para analizar si existe alguna ventaja evolutiva a favor de la cooperación y la especialización, en esta sección extenderemos el modelo causal propuesto en la sección \ref{sec:methodology} incorporando comportamientos incondicionalmente cooperativos, como el presentado en la sección \ref{sec:introduction}, incluyendo comportamientos desertores (que reciben el beneficio de la cooperación sin aportar al mismo). }%
%
\en{Cooperation in a non-randomly distributed population induces groups formation and therefore a type of group-level selection that acts in parallel to individual-level selection. }%
\es{La cooperación en poblaciones distribuidas no aleatoriamente induce la formación de grupos y por lo tanto un tipo de selección a nivel grupal que actúa de forma paralela a la selección de nivel individual. }%
%
\en{Darwin has already discussed this topic. }
\es{Darwin ya discutió sobre este tema. }
%
\begin{quotation}
It must not be forgotten that although a high standard of morality gives but a slight or no advantage to each individual (...), an advancement in the standard of morality will certainly given an immense advantage to one tribe over another.
\end{quotation}
% \en{This modification alone has a fundamental evolutionary consequence: the formation of groups and thus the emergence of a new evolutionary entity. }%
% \es{Esta sola modificación tiene una consecuencia evolutiva fundamental: la formación de grupos y por lo tanto la aparición de una nueva entidad evolutiva. }
%
\en{In other words, while defecting individuals may have a relative advantage over cooperating individuals, groups may be favored by mutual cooperation. }%
\es{En otras palabras, si bien los individuos desertores puede tener una ventaja relativa sobre los individuos cooperadores, los grupos pueden verse favorecidos por la mututa cooperación. }%
%
\en{The result of both selection processes, known as multilevel selection, is the property of interest. }%
\es{El resultado de ambos procesos de selección, conocido como selección multinivel, es la propiedad de nuestro interés. }%

\subsection{\en{Extended causal model}\es{Modelo causal extendido}}

\en{Individuals are spatially distributed and can interact only with members of the same region. }%
\es{Los individuos están distribuidos en el espacio y pueden interactuar solamente con los miembros de la misma región. }%

@@

\en{Then, there are $N+1$ types of possible regions, from $n=0$ (all defectors) to $n=N$ (all cooperators). }%
\es{Luego, hay $N+1$ tipos de regiones posibles, desde $n=0$ (todos desertores) hasta $n=N$ (todos cooperadores). }%

@@

\en{and the strategy $\Ee_i$ they use to allocate resources $f(\Ee,\Aa)$ (equation \ref{eq:familia_de_aptitudes}). }%
\es{y la estrategia $\Ee_i$ que utilizan para alocar recursos $f(\Ee,\Aa)$ (ecuación~\ref{eq:familia_de_aptitudes}). }%
\en{By combinatorics, it will be more likely a priori that individuals will inhabit mixed regions. }%
\es{Por combinatoria, será más probable a priori que los individuos habiten regiones mixtas. }%
P(i) = \frac{1}{N} \mathcal{B}(\texttt{region}(i)|N,0.5) 
\en{where the binomial distribution $\mathcal{B}$ gives more weight to individuals from mixed regions, the parameter $0.5$ indicates a uniform prior between cooperating and defecting social behaviors, and the factor $\frac{1}{N}$ indicates that all individuals in the same region start with the same initial resources. }%
\es{donde la distribución binomial $\mathcal{B}$ otorga más peso a los individuos de las regiones mixtas, el parámetro $0.5$ indica un prior uniforme entre los comportamientos sociales cooperador y desertor, y el factor $\frac{1}{N}$ indica que todos los individuos de la misma región empiezan con los mismos recursos iniciales. }%
\en{As we have seen in the section \ref{sec:methodology}, the probability of the environmental states is, }%
\es{Como hemos visto en la sección \ref{sec:methodology}, la probabilidad de los estados ambientales es, }%

@@

\en{Finally, the groups are made up of members from each region, $P(g|i)=\mathbb{I}(\texttt{region}(i) = g)$, the indexing function which is $1$ when individuals belong to region $g$ and $0$ otherwise. }%

@@

\en{The evolutionary problem we are interested in is the selection of cooperative individuals given the environment within each group (level 1), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)$, the selection of groups given the environment (level 2), $P(g^T|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$,  and selection of cooperative individuals given the environment for all groups (multilevel), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$. }%
\es{El problema evolutivo que nos interesa es la selección de los individuos cooperadores dado el ambiente dentro de cada grupo (nivel 1), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)$, la selección de los grupos dado el ambiente (nivel 2), $P(g^T|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$, y la selección de los individuos cooperadores dado el ambiente integrando todos los grupos (multinivel), $P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})$. }%
\en{The multilevel selection is obtained by integrating the product of level 1 and 2 selections, }%
\es{La selección multinivel se obtiene integrando el producto de las selecciones de nivel 1 y 2, }%
\begin{equation}\label{eq:posterior_multinivel}
\underbrace{P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})}_{\text{\en{Multilevel selection}\es{Selección multinivel}}} = \sum_{g=0}^N \underbrace{P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)}_{\text{\en{Level 1 selection}\es{Selección de nivel 1}}} \cdot \underbrace{P(g^T|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1})}_{\text{\en{Level 2 selection}\es{Selección de nivel 2}}}
\en{where the posterior of level 1 selection is obtained by integrating the probability of all cooperating individuals that are part of the group $g$, }%
\es{donde el posterior de la selección de nivel 1 se obtiene integrando la probabilidad de todos los individuos cooperadores que forman parte del grupo $g$, }%
\begin{equation}\label{eq:posterior_nivel_1}
P(\texttt{coop}(i^T)|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g) &= \sum_{j=1}^M P(I^T=j|\vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T-1}, g)\mathbb{I}(\texttt{coop}(j))
\en{where}\es{donde} (\en{all steps are described in detail in the appendix}
\es{en el anexo resolvemos todos los pasos detalladamente})
\en{where $r = \texttt{region}(k)$ is the region to which individual $k$ belongs, $\texttt{partners}(r)$ is the set of cooperating individuals in region $r$, and $c$ is a cooperative individual belonging to that specific region, $c \in \texttt{partners}(r)$. }%s
\es{donde $r = \texttt{region}(k)$ es la región a la que pertenece el individuo $k$, $\texttt{\en{partners}\es{socios}}(r)$ es el conjunto de individuos cooperadores de la región $r$, y $c$ es un individuo cooperador de esa región, $c \in \texttt{\en{partners}\es{socios}}(r)$. }%

\subsection{\en{The multiplicative nature of evolutionary and probability theories}\es{La naturaleza multiplicativa de las teorías de la evolución y la probabilidad}}

\en{Even if our causal model does not assume any kind of process to update resources, the result of the inference is proportional to the resources obtained through a multiplicative process. }%
\es{Incluso si nuestro modelo causal no presupone ningún tipo de proceso para actualizar los recursos, el resultado de la inferencia es proporcional a los recursos obtenidos a través de un proceso multiplicativo. }%
% \en{The proposed causal model does not require assuming that the resource updating process is multiplicative, as proposed by Ole Peters for his game. }%
% \es{El modelo causal propuesto no requiere suponer que el proceso de actualización de los recursos sea multiplicativo, como propone Ole Peters para su juego. }%
% %
\en{All the assumptions of our model are made explicit in the Bayesian network in Figure~\ref{fig:multilevel_model}. }
\es{Todos los supuestos de nuestro modelo se encuentran explicitados en la red bayesiana de la figura~\ref{fig:multilevel_model}. }%
%
\en{Individuals are affected just by the environment and by the social behaviors of cooperation and defection of their context. }%
\es{Los individuos se ven afectados solamente por el ambiente y por los comportamientos sociales de cooperación y deserción de su contexto. }%
%
\en{However, our probabilistic causal model and the multiplicative process discussed in the introduction are equivalent. }%
\es{Sin embargo, nuestro modelo causal probabilístico y el proceso multiplicativo analizado en la introducción son equivalentes. }%
%  
% % Parrafo
% 
% \en{Without the need to presuppose a multiplicative process, the posteriors of individuals in our causal model are proportional to the trajectory of resources arising from the game proposed by Ole Peters. }%
% \es{Sin necesidad de presuponer un proceso multiplicativo, los posteriors de los individuos en nuestro modelo causal son proporcionales a la trayectoria de los recursos que surgen del juego propuesto por Ole Peters. }%
%
\en{The posterior of individuals from the entirely defecting region is, }%
\es{El posterior de los individuos de la región enteramente desertora es, }%
%
\begin{equation}
\begin{split}
P(i^{T+1} | \vec{\Aa}^{\,1}, \dots, \vec{\Aa}^{\,T}) & \overset{r=0}{=}  P(i^{T+1})  \prod_{t=1}^{T} P(i^{T+1}|\vec{\Aa}^{\,t}) \\
& \propto P(i^{T+1}) \prod_{t=1}^{T} \Ee_k^{\Aa_k} (1-\Ee_k)^{1-\Aa_k}   \propto P(i^{T+1}) \prod_{t=1}^{T} R f(\Ee_i,\Aa_{i}^{t})
\end{split}
\end{equation}
%
\en{where $r = \texttt{region}(i^{T+1}) = 0$ denotes the region without cooperators, $f$ is the fitness function that we define in the equation \ref{eq:familia_de_aptitudes}. }%
\es{donde $r = \texttt{region}(i^{T+1}) = 0$ indica la región sin cooperadores, $f$ es la familia de aptitudes que definimos en la ecuación \ref{eq:familia_de_aptitudes}. }%
%
\en{The first proportional is valid because of the normalization constant of the probability distribution $P(i^t|\vec{\Aa}^{\,t})$. }%
\es{El primer proporcional vale por constante de normalización de la distribución de probabilidad $ P(i^t|\vec{\Aa}^{\,t})$. }%
%
\en{And the second proportional simply includes the factor $R$, so that the fitness function matches the one proposed by Lewontin-Cohen, when $R=2.2$, or the one proposed by Peters, when $R=2.1$. }%
\es{Y el segundo proporcional simplemente incluye el factor $R$ para que la función de aptitudes coincida con la prouesta por Lewontin-Cohen, cuando $R=2.2$, o la propuesta por Peters, $R=2.1$. }%
%
\en{When the prior $P(i^{T+1})$ represents the initial resources $\omega(0)$, then the posterior of the individual defectors in entirely defecting groups is proportional to the growth of the individual resources studied in the introduction section. }%
\es{Si consideramos que el prior $P(i^{T+1})$ representa los recursos inciales $\omega(0)$, luego el posterior de los individuos desertores en grupos enteramenete desertoras es proporcional al crecimiento de los recursos individuales estudiada en la introducción. }%

\en{The same is true for the posteriors of cooperating individuals. }%
\es{Lo mismo ocurre con los posteriors de los individuos cooperadores. }%
%
\en{First, the message sent by the individual variables to the social factor is proportional to the second node of the cooperative protocol (figure~\ref{fig:protocolo}), in which the previous resources $\omega_i(t)$ are updated by the product of the fitness function. }%
\es{Primero, el mensaje que envían las variables individuos al factor social es proprocional al segundo nodo de del protocolo cooperativo (figura~\ref{fig:protocolo}), en el que los recursos previos $\omega_i(t)$ mediante el producto de la función de fitness. }%
%
\begin{equation}
\begin{split}
m_{I^t \rightarrow P(I^{t+1}|I^t)}(i^t) & = m_{P(I^t|\vec{\A}^{\,t}) \rightarrow I^t}(i^t) \, m_{P(I^t|I^{t-1}) \rightarrow I^t }(i^t) \\
& = P(i^t| \vec{\Aa}^{\,t}) \, P(i^t, \vec{\Aa}^{\,t}, \dots, \vec{\Aa}^{\,t}) \\
& \propto R\,f(\,\Ee_i,\Aa_{i}^{t}) \omega(t-1)
\end{split}
\end{equation}
%
%donde la marginal $P(i^t, \vec{\Aa}^{\,t}, \dots, \vec{\Aa}^{\,t})$ es proporcional al posterior $P(i^t| \vec{\Aa}^{\,t}, \dots, \vec{\Aa}^{\,t})$.
%
\en{Again, the proportionality holds by construction of the probability distribution $P(i^t|\vec{\Aa}^{\,t})$. }%
\es{Nuevamente, la prorporcionalidad vale por construcción de la distribución de probabilidad $P(i^t|\vec{\Aa}^{\,t})$. }%
%
\en{And the message that the social factor sends to the individual variable of the following time is proprotional to the individual resources after pooling and sharing, }%
\es{Y el mensaje que envía el factor social a la variable individuo del siguiente tiempo es proprocional a los recursos individuales luego de la redistribución del fondo común, }%
%
\begin{equation}
\begin{split}
m_{P(I^{t+1}|I^{t}) \rightarrow I^{t+1} }(i^{t+1}) & = \sum_{i^t} P(i^{t+1}|i^t) \, P(\vec{\Aa}^{\,t}) \, P(i^t|\vec{\Aa}^{\,t}) \,  m_{P(I^t|I^{t-1}) \rightarrow I^t }(i^t) \\
%& \propto \sum_{i^t} P(i^{t+1}|i^t) \, P(i^t|\vec{\Aa}^{\,t}) \,  \text{Prior}(i^t) \\
= P(I^{t+1}, \vec{\Aa}^{\,1} , \dots, \vec{\Aa}^{\,T} ) & \propto \sum_{i^t} P(i^{t+1}|i^t) \, f_{\Ee_i}(\Aa_{i}^{t}) \omega_i(t)  = \omega_i(t+1)
\end{split}
\end{equation}
%
\en{where $P(i^{t+1}|i^t) = \frac{1}{N}$ when individuals belong to the same region $\texttt{region}(i^{t+1}) = \texttt{region}(i^{t})$. }%
\es{donde $P(i^{t+1}|i^t) = \frac{1}{N}$ cuando los individuos pertenecen a la misma región $\texttt{region}(i^{t+1}) = \texttt{region}(i^{t})$. }%
%
\en{This message generalizes the last node of the cooperative protocol (figure~\ref{fig:protocolo}). }%
\es{Este mensaje generaliza el último nodo del protocolo cooperativo (figura~\ref{fig:protocolo}). }%

\en{In the introduction we have not analyzed the resource trajectory of defectors in cooperative groups, but in general, the posterior of individuals is no more than the proportion of resources they manage. }%
\es{En la introducción no hemos analizado la trayectoria de los recursos de los desertores en grupos cooperativos, pero en general, el posterior de los individuos no es más que la proporción de recursos que maneja. }%
%
\begin{equation}
P(k| \vec{\Aa}^{\,1}, \dots,  \vec{\Aa}^{\,T}) = \frac{\omega_k(T)}{\sum_j \omega_j(T)}
\end{equation}
%
\en{The proportionality between the posterior probability of individuals and the trajectories of resources studied in the introduction section allows us to work with both expressions interchangeably. }%
\es{La proporcionalidad entre la probabilidad a posteriori de los individuos y las trayectorias de los recursos estudiada en la instroducción nos permite trabajar indistintamanete con una u otra expresión. }%

\en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes~\cite{dempster1955-geometricMean, denBoer1968-spreadingRisk}. }%
\es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos~\cite{dempster1955-geometricMean, denBoer1968-spreadingRisk}. }% 
%
\en{This coincidence is an additional support to the hypothesis of isomorphism between evolutionary and probability theories, previously identified between their fundamental equations: the Bayes theorem and the replicator dynamic~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
\es{Esta coincidecia es un apoyo adicional a la hipótesis de isomorfimo entre las teorías de la evolutivas y probabilisticas, previamente identificado entre sus ecuaciones fundamentales: el teorema de bayes y el replicator dynamic~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%

\subsection{\en{Commons dilemma}\es{Dilema de los bienes comunes}}

%En vez de utilizar una matriz de pagos del dilema del prisionero para representar un proceso de bienes comunes, en nuestro trabajo definimos el proceso de bienes comunes a partir del cual podemos derivar la matriz de pagos asociada.
%
\en{If our causal model had a payoff structure isomorphic to the prisoner's dilemma, then defectors would have a higher growth rate than cooperators. }%
\es{Si nuestro modelo causal tuviera una estrucutra de pagos isomorfa al dilema del prisioner, entonces los desertores tendrían una tasa de crecimiento mayor a lo cooperadores. }%
%
\en{However, the first defector from an entirely cooperative group obtains a lower growth rate than before defecting. }%
\es{Sin embargo, el primer desertor de un grupo enteramente cooperadora obtiene una tasa de crecimiento menor a la que tenía antes de desertar. }% 
%
\en{In the figure~\ref{fig:ergodicity_desertion} we can observe the trajectories of the resources, equivalent to the proportional posterior (see previous section), of the individuals that are in a group of size 100. }%
\es{En la figura~\ref{fig:ergodicity_desertion} podemos observar las trayectoria de los recursos, equivalente el posterior proporcional (ver sección anterior), de los individuos que están en un grupos de tamaño 100. }%
%
\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.5\textwidth}
    \includegraphics[width=\linewidth]{figures/pdf/ergodicity_desertion.pdf}
    \end{subfigure}
    \caption{
    \en{The colors represent groups of size 100 with 0, 1 and 2 defectors. }%
    \es{Los colores representan los grupos de tamaño 100 con 0, 1 y 2 desertores. }%
    %
    \en{The curves of the individual defectors are those at the top of each of the groups. }%
    \es{Las curvas de los individuos desertores son las que están arriba en cada uno de los grupos. }%
    }
    \label{fig:ergodicity_desertion}
\end{figure}
%
\en{The resources of the first individual defector (blue group with 1 defector), is below the resources of the individuals in the fully cooperative group (green group with 0 defectors). }%
\es{Los recursos del primer individuo desertor (grupo azul con 1 desertor), está por debajo de los recursos de los individuos del grupo enteramente cooperador (grupo verde con 0 desertores). }%
%
\en{The reduction in resources occurs even for the second individual who changes from cooperative to a defective behavior. }%
\es{La reducción de recursos le ocurre incluso al segundo individuo que cambia de comportamiento cooperador a desertor. }%

\en{Defecting, instead of increasing the growth rate of the defective individuals, reduces it. }%
\es{Desertar, en vez de aumentar la tasa de crecimiento de los individuos desertores, la reduce. }%
%
\en{In other words, the commons does not have the structure of the prisoner's dilemma, as is usually claimed in the literature. }%
\es{Es decir, los bienes comunes no tiene la estructura del dilema del prisionero como habitualmente se afirma en la literatura. }%
%
\en{Let us calculate the payoff matrix arising from the cooperative causal model. }%
\es{Calculemos la matriz de pagos que surge del modelo causal cooperativo. }%
%
\en{We want to estimate the temporal growth rate of the posteriors, }%
\es{Queremos estimar la tasa de crecimiento temporal de los posteriors, }%
%
\begin{equation}
\frac{P(k|\vec{\Aa}^{\,1}, \dots , \vec{\Aa}^{\,T})}{P(k)} \approx g(k|\vec{\Ee},p,N)^T
\end{equation}
%
\en{where the approximation is an equality when time tends to infinity, $\lim_{T \rightarrow \infty}$. }%
\es{donde la aproximación es una igualdad cuando el tiempo tiende a infinito, $\lim_{T \rightarrow \infty}$. }%
%
\en{The growth rate depends on the hyperparameters of the model, $\vec{\Ee}$, $p$ and $N$. }%
\es{La tasa de crecimiento depende de los hyperparámtros del modelo, $\vec{\Ee}$, $p$ y $N$. }%
%
\en{Here we will only consider models in which all individuals have the same strategy $\Ee$. }%
\es{Aquí vamos a considerar únicamente los modelos en el que todos los individuos tienen la misma estrategia $\Ee$. }%
%
\en{We will solve this problem by cases, depending on whether the behavior is cooperative, $g_C$, or defective, $g_D$, for different social contexts $g_{[\cdot]}^n(k|\cdot)$, where $n$ represents the number of cooperators with which the individual $k$ interacts (excluding $k$). }%
\es{Resolveremos este problema por casos, según el comportamiento sea cooperativo, $g_C$, o desertivos, $g_D$, para diferente contextos sociales $g_{[\cdot]}^n(k|\cdot)$, donde $n$ representa la cantidad de cooperadores con lo que interactúa el individuo $k$ (excluyendo a $k$). }%
% 
\en{The growth rate of mutual defection is, }%
\es{La tasa de crecimiento de la deserción mutua es, }%
%
\begin{equation}
\begin{split}
\lim_{T \rightarrow \infty} g_D^0(k|\Ee,p,N)^T &= \prod_{t=1}^T P(k|\vec{\Aa}^{\,T+1}) \propto \prod_{t=1}^T \Ee^{\Aa^t_k} (1-\Ee)^{1-\Aa^t_k} \\
\lim_{T \rightarrow \infty} g_D^0(k|\Ee,p,N) &\propto \Big( \prod_{t=1}^T \Ee^{\Aa^t_k} (1-\Ee)^{1-\Aa^t_k} \Big) ^{1/T} = \Ee^{p} (1-\Ee) ^{1-p}
\end{split}
\end{equation}
%
\en{which is independent of the size $N$. }%
\es{que es independiete del tamaño $N$. }%
%
\en{The proportional is valid because of the normalization constant of the probability distribution $P(i^t|\vec{\Aa}^{\,t})$. }%
\es{El proporcional vale por constante de normalización de la distribución de probabilidad $ P(i^t|\vec{\Aa}^{\,t})$. }%
%
\en{The growth rate of a strategy $\Ee= 0.71 \approx 1.5/2.1$ in an environment with $p=0.5$ is $ 0.71^{1/2}\cdot0.29^{1/2} \approx 0.452$. }%
\es{La tasa de crecimiento de un estrategia $\Ee= 0.71 \approx 1.5/2.1$ en una ambiente con $p=0.5$ es $ 0.71^{1/2}\cdot0.29^{1/2} \approx 0.452$. }%
